\documentclass[10pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsfonts}
\usepackage{booktabs}
\usepackage{enumerate}
\usepackage{hyperref}

\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\ind}[1]{\mathbb{1}\!\left[#1\right]}
\newcommand{\Tok}{\mathcal{T}}
\newcommand{\Dir}{\mathrm{Dir}}
\newcommand{\GEM}{\mathrm{GEM}}
\newcommand{\DP}{\mathrm{DP}}
\newcommand{\NIW}{\mathrm{NIW}}
\newcommand{\NB}{\mathrm{NB}}
\newcommand{\Norm}{\mathcal{N}}
\newcommand{\sil}{\langle\mathrm{sil}\rangle}
\newcommand{\noise}{\langle\mathrm{noise}\rangle}
\newcommand{\softmax}{\mathrm{softmax}}
\newcommand{\LN}{\mathrm{LN}}
\newcommand{\FFN}{\mathrm{FFN}}
\newcommand{\MHA}{\mathrm{MHA}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

\title{\textbf{kashi v2: Mathematical Specification of the Textless\\ Syllable-Level Transcription Pipeline}}
\author{Companion to \emph{Textless Syllable Alignment of Japanese Songs} (CS 541 report)\\ and to \texttt{ROADMAP.md}}
\date{2026-07-09 \quad (planning document; no implementation yet)}

\begin{document}
\maketitle

\noindent\textbf{Scope.} Complete mathematical definition of every model in the planned pipeline: vocal separation, frame features (wav2vec2, contrastive projection, voicing/TDA), Model~1 (boundary scorer, corrected), Model~2 (syllable classifier, segment and frame variants), the phonetic kernel, the semi-Markov segmental decoder that joins them, the sticky HDP-HMM/HSMM used for unsupervised structure, the alignment-free dataset-cleaning procedure, unsupervised objectives, and evaluation metrics. Design constraints throughout: (i) \emph{textless} --- no transcript is available or used at inference; (ii) \emph{no forced alignment anywhere}, including dataset cleaning (Remark~\ref{rem:nofa}); (iii) deep networks supply local evidence, while segmentation, duration, and label structure are explicit probabilistic components rather than black-box behavior.

\section{Notation and problem statement}\label{sec:notation}

Notation follows the report; collisions in the report are resolved as footnoted.

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Symbol & Meaning & Report correspondence \\
\midrule
$a \in \R^{N_a}$ & mono waveform, $f_s = 16\,000$ Hz & (audio file) \\
$h = 0.02$ s & frame hop; $T = \lfloor N_a/(f_s h)\rfloor$ frames & 20-ms interval \\
$x_t \in \R^{768}$ & wav2vec2 vector of frame $t \in \{1,\dots,T\}$ & $x_t$ \\
$\tilde{x}_t \in \R^{d}$ & final frame feature (\S\ref{sec:features}) & --- (new) \\
$\Tok$, $|\Tok| = 110$ & token set: 109 hiragana + $\sil$ & $T$ (set)\footnotemark \\
$y_t \in [0,1]$ & Model-1 per-frame break score & $y_t$ \\
$\sigma_k = (t_k, s_k, e_k)$ & segment $k$: token $t_k \in \Tok$, start $s_k$, end $e_k$ (s) & $s_k = (t_k, s_k, e_k)$\footnotemark \\
$\sigma = (\sigma_1,\dots,\sigma_K)$ & full segmentation; $s_1 = 0$, $s_{k+1} = e_k$, $e_K = Th$ & $\{s_k\}_{k=0}^{K}$ \\
\bottomrule
\end{tabular}
\end{center}
\footnotetext[1]{The report uses $T$ for both the token set and the sequence length in $f(x_1,\dots,x_T)$; we write $\Tok$ for the set and keep $T$ for length.}
\footnotetext[2]{The report overloads $s_k$ as both the segment and its start time; we write $\sigma_k$ for the segment.}

\paragraph{Task.} Learn $F: a \mapsto \sigma$ from a corpus $\mathcal{D} = \{(a^{(i)}, \hat{\sigma}^{(i)})\}_{i=1}^{n}$ whose label \emph{times} are corrupted,
\begin{equation}
\hat{s}_k = s_k^\star + \varepsilon_k, \qquad |\varepsilon_k| \le \varepsilon_{\max} \approx 50\text{ ms} \;(= 2.5\text{ frames}),
\label{eq:jitter}
\end{equation}
whose \emph{sequences} $(t_1,\dots,t_K)$ are reliable, and whose $\sil$ segments may contain unlabeled vocal events (breaths). At inference only $a$ is given.

\begin{samepage}
\begin{list}{}{\leftmargin=1.5em}
\item \textbf{Remark 1 (exclusion of forced alignment).}\label{rem:nofa}
Forced alignment solves, for a \emph{given} token sequence $w_{1:M}$,
$\hat{A} = \argmax_{A \in \mathcal{A}(w_{1:M}, T)} p(x_{1:T}, A \mid w_{1:M})$
over monotone alignments $\mathcal{A}$. Every procedure below is required \emph{not} to solve this problem or any restriction of it: no stage conditions a time-inference on a token sequence. Training may use the labeled sequences as supervision targets (all supervised learning does); what is excluded is deriving \emph{timestamps} by aligning a transcript --- at inference (no transcript exists; singers go off script) and during dataset cleaning (\S\ref{sec:cleaning} uses only sequence-free local corrections).
\end{list}
\end{samepage}

\paragraph{Pipeline operator.} $F = W \circ D \circ \Phi \circ V$: separation $V$ (\S\ref{sec:sep}), frame features $\Phi$ (\S\ref{sec:features}), segmental decode $D$ (\S\ref{sec:decoder}) consuming the model scores of \S\S\ref{sec:model1}--\ref{sec:model2}, subtitle writing $W$ (exact; not modeled). Each stage is a swappable component; this document specifies the default mathematics of each slot.

\section{Stage $V$: vocal separation (off the shelf)}\label{sec:sep}

With $X = \mathrm{STFT}(a) \in \mathbb{C}^{F \times N}$, a separator computes a vocal estimate $\hat{a} = \mathrm{iSTFT}\!\left(M_\theta(X) \odot X\right)$ where $M_\theta \in [0,1]^{F\times N}$ (or complex-valued) is produced by a pretrained network (default: Mel-Band RoFormer checkpoint; alternates: Demucs-family time-domain models, identity). $V$ is treated as a fixed operator: its parameters are never trained here, and the report verified label preservation on the corpus. All downstream stages see only $\hat{a}$, resampled to 16 kHz mono.

\section{Stage $\Phi$: frame representation}\label{sec:features}

\subsection{wav2vec2 backbone}
A pretrained wav2vec2/HuBERT encoder maps $\hat a$ to latents $z_\tau$ (convolutional encoder, 20-ms stride, $\approx$25-ms receptive field) and context vectors
\begin{equation}
x_t = \mathrm{TransformerEncoder}(z)_t \in \R^{768}, \qquad t = 1,\dots,T .
\end{equation}
The checkpoint is a config choice (default candidates: \texttt{rinna/japanese-hubert-base}, \texttt{rinna/japanese-wav2vec2-base}; legacy: \texttt{facebook/wav2vec2-base}). The backbone is frozen in all default configurations.

\subsection{Learned projection (contrastive; optional)}\label{sec:proj}
$g_w: \R^{768} \to \R^{128}$,
\begin{equation}
g_w(x) = \frac{W_2\,\phi(W_1 x + b_1) + b_2}{\lVert W_2\,\phi(W_1 x + b_1) + b_2 \rVert_2},
\qquad W_1 \in \R^{256\times768},\; W_2 \in \R^{128\times256},\; \phi = \mathrm{GELU},
\end{equation}
trained by the objectives of \S\ref{sec:unsup} on cached features (backbone frozen). When enabled, $g_w(x_t)$ replaces $x_t$ everywhere downstream (deep-kernel view: $g_w$ \emph{is} the learned similarity).

\subsection{Voicing score $v_t$ (periodicity)}\label{sec:voicing}
Let $w_t \in \R^{L_w}$ be the waveform window of $46$ ms centered on frame $t$, mean-removed. Normalized autocorrelation and voicing:
\begin{equation}
\rho_t(\ell) = \frac{\sum_{n} w_t[n]\, w_t[n+\ell]}{\sqrt{\sum_n w_t[n]^2 \sum_n w_t[n+\ell]^2}},
\qquad
v_t = \max_{\ell \in [f_s/f_{\max},\, f_s/f_{\min}]} \rho_t(\ell),
\quad (f_{\min}, f_{\max}) = (70, 500)\,\mathrm{Hz}.
\label{eq:autocorr}
\end{equation}
Voiced morae (vowels; voiced consonants \emph{ga, wo}, \dots) are quasi-periodic $\Rightarrow v_t \to 1$; breath/silence are aperiodic $\Rightarrow v_t$ small. \emph{TDA variant} (interchangeable estimator of the same quantity): Takens delay embedding $y_n = (w_t[n], w_t[n+\tau], w_t[n+2\tau]) \in \R^3$ with $\tau = f_s/(2 f_{\max})$; Vietoris--Rips filtration on $\{y_n\}$; with $\mathrm{PD}_1$ the $H_1$ persistence diagram,
\begin{equation}
v_t^{\mathrm{TDA}} = \frac{\max_{(b,d) \in \mathrm{PD}_1} (d-b)}{\mathrm{diam}\{y_n\}} \in [0,1],
\end{equation}
i.e.\ the longest normalized loop lifetime (periodic $\Leftrightarrow$ closed orbit $\Leftrightarrow$ persistent $H_1$ class). Default is \eqref{eq:autocorr} (cheap, no dependency); the TDA estimator is a config alternate benchmarked once.

\subsection{Final frame feature}
\begin{equation}
\tilde{x}_t = \big[\, g_w(x_t) \text{ (or } x_t\text{)};\; v_t;\; e_t \,\big], \qquad e_t = \log \mathrm{RMS}(w_t),
\end{equation}
so $d \in \{770, 130\}$ depending on projection use. The scalar channels give every downstream model explicit, interpretable voicing/energy evidence (targets the \emph{ka/ga}, \emph{wo/o} confusion axis and breath detection).

\section{Model 1: boundary scorer}\label{sec:model1}

\subsection{Architecture (corrected)}
Per-frame boundary logits from a temporal transformer:
\begin{align}
u_t^{(0)} &= W_{\mathrm{in}} \tilde{x}_t + p_t, \qquad W_{\mathrm{in}} \in \R^{d_m \times d},\; d_m = 256,\\
p_t[2i] &= \sin\!\big(t / 10000^{2i/d_m}\big), \qquad p_t[2i{+}1] = \cos\!\big(t / 10000^{2i/d_m}\big),\\
\intertext{for layers $\ell = 1,\dots,L$ ($L=2$), with $U = [u_1;\dots;u_T]$ processed \emph{as a length-$T$ sequence}:}
\mathrm{head}_i(U) &= \softmax\!\Big(\tfrac{(U W_i^Q)(U W_i^K)^\top}{\sqrt{d_k}}\Big)\, U W_i^V, \qquad i = 1,\dots,H,\; H = 8,\; d_k = d_m/H,\\
\MHA(U) &= [\mathrm{head}_1;\dots;\mathrm{head}_H]\, W^{O},\\
U' &= \LN\big(U + \MHA(U)\big), \qquad
U^{(\ell)} = \LN\big(U' + \FFN(U')\big),\\
\FFN(u) &= W_2^{\mathrm{ff}}\,\mathrm{ReLU}(W_1^{\mathrm{ff}} u + b_1) + b_2, \qquad W_1^{\mathrm{ff}} \in \R^{512 \times d_m},\\
\beta_t &= w_{\mathrm{out}}^\top u_t^{(L)}, \qquad y_t = \sigma(\beta_t) = \big(1+e^{-\beta_t}\big)^{-1}.
\end{align}
Attention masking limits context to $\pm W_a = \pm 100$ frames ($\pm 2$ s; boundaries are local evidence). \emph{Corrections vs.\ the recovered legacy implementation} (\texttt{google drive/Code/Transformer\_Wave2Vec\_AAAAA.ipynb}): (i) attention must run over the \emph{time} axis (the legacy code omitted \texttt{batch\_first=True}, so self-attention saw length-1 sequences --- no temporal context); (ii) positional encodings $p_t$ were absent; (iii) label expansion must be edge-guarded (below).

\subsection{Label construction and losses}
Break frames from labels: $\hat{b}_j = \mathrm{frame}(\hat{e}_{k_j})$ for internal segment ends, $j = 1,\dots,J$, excluding excluded rows. Two supervised objectives (both implemented; selected by the metric in \S\ref{sec:metrics}):

\textbf{(a) Soft-kernel BCE (report's method, regularized).} Targets by edge-guarded kernel placement,
\begin{equation}
\tilde{y}_\tau = \min\Big(1, \sum_{j=1}^J \kappa(\tau - \hat{b}_j)\Big), \qquad
\kappa(-2{:}3) = (0.2,\, 0.35,\, 1,\, 1,\, 0.8,\, 0.25),\; \kappa = 0 \text{ otherwise},
\end{equation}
\begin{equation}
\mathcal{L}_{\mathrm{soft}} = -\tfrac{1}{T}\textstyle\sum_{\tau=1}^{T} m_\tau \big[ w_+ \tilde{y}_\tau \log y_\tau + (1-\tilde{y}_\tau)\log(1-y_\tau) \big],
\end{equation}
$m_\tau \in \{0,1\}$ masking excluded/padded frames, $w_+ = $ (\#neg/\#pos) on unexpanded labels.

\textbf{(b) Latent-offset marginalization (principled jitter model; new).} Model \eqref{eq:jitter} directly: the true break is $b_j = \hat{b}_j + \delta_j$ with $\delta_j \overset{\text{iid}}{\sim} \pi$ on $\{-\Delta,\dots,\Delta\}$, $\Delta = 3$ frames $\ge \varepsilon_{\max}$, $\pi$ triangular. With per-frame Bernoulli emissions $y_\tau$ and disjoint windows $W_j = [\hat b_j - \Delta,\, \hat b_j + \Delta]$ (guaranteed by minimum inter-break distance; ties broken by window truncation),
\begin{equation}
\mathcal{L}_{\mathrm{lom}} = -\sum_{j=1}^{J} \log \sum_{\delta = -\Delta}^{\Delta} \pi(\delta)\; y_{\hat b_j + \delta} \!\!\prod_{\substack{\tau \in W_j \\ \tau \ne \hat b_j + \delta}}\!\! (1 - y_\tau)
\;-\; w_-\!\!\sum_{\tau \notin \cup_j W_j}\!\! m_\tau \log (1 - y_\tau).
\label{eq:lom}
\end{equation}
Equation~\eqref{eq:lom} is the exact marginal likelihood of the noisy annotation under the jitter model --- the model is never penalized for placing the break anywhere inside the tolerance window, which is what \eqref{eq:jitter} says we actually know. It involves no sequence and no alignment; each window is an independent local latent variable.

\subsection{Boundary decoding (standalone use)}
\begin{equation}
\hat{B} = \{ t : y_t \ge \theta,\;\; y_t = \max_{|t'-t| \le \nu} y_{t'} \}, \qquad \theta = 0.45,\; \nu = 3 \text{ frames (NMS)}.
\end{equation}
In the full pipeline Model 1 is \emph{not} thresholded; its logits enter the decoder as boundary potentials (\S\ref{sec:decoder}).

\section{Model 2: syllable classifier and phonetic kernel}\label{sec:model2}

\subsection{Segment variant (report architecture, retained)}
For a span $[s, e)$ with frames $\tilde x_{s:e}$, a 2-layer LSTM (hidden $n_h = 144$, dropout $0.5$ between layers) runs
\begin{align}
i_\tau &= \sigma(W_{xi}\tilde x_\tau + W_{hi} h_{\tau-1} + b_i), &
f_\tau &= \sigma(W_{xf}\tilde x_\tau + W_{hf} h_{\tau-1} + b_f), \nonumber\\
g_\tau &= \tanh(W_{xg}\tilde x_\tau + W_{hg} h_{\tau-1} + b_g), &
o_\tau &= \sigma(W_{xo}\tilde x_\tau + W_{ho} h_{\tau-1} + b_o), \\
c_\tau &= f_\tau \odot c_{\tau-1} + i_\tau \odot g_\tau, &
h_\tau &= o_\tau \odot \tanh(c_\tau), \nonumber
\end{align}
and classifies from the final hidden state of the top layer:
\begin{equation}
p_{\mathrm{M2}}(u \mid \tilde x_{s:e}) = \softmax\big(W_c\, h^{(2)}_{e-1} + b_c\big)_u, \qquad W_c \in \R^{110 \times n_h}.
\end{equation}
Parameter names (\texttt{lstm}, \texttt{fc}) and class order are frozen to load the legacy checkpoint (test acc.\ $0.536$) as initialization.

\subsection{Frame variant (new; enables exact decoding)}\label{sec:m2f}
A per-frame posterior over the same classes,
\begin{equation}
p_{\mathrm{f}}(u \mid \tilde x_t) = \softmax\big(W_f\, \phi_f(\tilde x_t) + b_f\big)_u,
\end{equation}
with $\phi_f$ a 1-hidden-layer MLP ($256$, GELU) or identity. Trained on frames inside labeled segments (target = the segment's token; frames inside $\noise$ spans excluded; $\sil$ is an ordinary class), with the phonetic soft targets of \S\ref{sec:kernel}. Its role: additive segment scores $\sum_{t \in [s,e)} \log p_{\mathrm{f}}(u \mid \tilde x_t)$ decompose over frames, allowing the semi-Markov DP of \S\ref{sec:decoder} to search \emph{all} segmentations exactly with cumulative sums; the LSTM variant then rescores finalists.

\subsection{Phonetic kernel $k$ on $\Tok$}\label{sec:kernel}
Decompose each non-silence token by its romaji as $\delta(u) = (c(u), \gamma(u), v(u))$: consonant $c \in \mathcal{C} \cup \{\varnothing\}$, palatal glide $\gamma \in \{0,1\}$ (\emph{kya, sho}, \dots), vowel $v \in \{a,i,u,e,o,\varnothing\}$ ($\varnothing$ for the moraic nasal $n$). Consonant features: $\mathrm{voice}(c) \in \{0,1\}$, $\mathrm{place}(c) \in \{$labial, alveolar, palatal, velar, glottal$\}$, $\mathrm{manner}(c) \in \{$plosive, fricative, affricate, nasal, approximant$\}$. Vowel coordinates $\varphi(v) = (\text{height}, \text{backness})$: $a = (0, 0.5)$, $i = (1, 0)$, $u = (1, 1)$, $e = (0.5, 0)$, $o = (0.5, 1)$.
\begin{align}
k_C(c, c') &= \lambda_v \ind{\mathrm{voice} =} + \lambda_p \ind{\mathrm{place} =} + \lambda_m \ind{\mathrm{manner} =}, \quad (\lambda_v, \lambda_p, \lambda_m) = (0.2, 0.4, 0.4),\nonumber\\
&\quad k_C(\varnothing,\varnothing) = 1, \quad k_C(\varnothing, c) = 0.15;\\
k_V(v, v') &= 1 - \tfrac{1}{2}\lVert \varphi(v) - \varphi(v') \rVert_1, \qquad k_V(\varnothing,\varnothing) = 1,\; k_V(\varnothing, v) = 0;\\
k(u, u') &= \mu_C\, k_C + \mu_V\, k_V + \mu_\gamma \ind{\gamma = \gamma'}, \qquad (\mu_C, \mu_V, \mu_\gamma) = (0.45, 0.45, 0.10),
\end{align}
with overrides: $k(u,u) = 1$; $k(\sil, u) = \ind{u = \sil}$; modern-Japanese homophone pairs $\{(\text{\emph{wo}},\text{\emph{o}}), (\text{\emph{di/ji}}), (\text{\emph{du/zu}})\}$ set to $0.95$. Worked values (matching the report's confusion structure): $k(\text{ka},\text{ga}) = 0.45{\cdot}0.8 + 0.45 + 0.1 = 0.91$ (voicing bit only); $k(\text{ka},\text{ta}) = 0.82$ (place differs); $k(\text{ka},\text{ki}) = 0.66$ (vowel moves); $k(\text{ka},\text{n}) = 0.10$. The full $110{\times}110$ PSD-projected matrix $K$ is precomputed once ($K \leftarrow K - \lambda_{\min}(K) I$ then renormalized, if needed, to guarantee positive semidefiniteness).

\emph{One kernel, three uses:}
\begin{enumerate}
\item \textbf{Partial-credit loss} (both classifier variants). Soft targets with sharpening power $p = 4$, mass $\alpha = 0.1$:
\begin{equation}
q(u' \mid u) = (1-\alpha)\ind{u'=u} + \alpha \frac{k(u,u')^{p}\,\ind{u' \ne u}}{\sum_{u'' \ne u} k(u,u'')^{p}}, \qquad
\mathcal{L}_{\mathrm{cls}} = -\sum_{u'} q(u' \mid u^\star) \log p(u' \mid \cdot).
\end{equation}
\item \textbf{Soft contrastive negatives} (\S\ref{sec:unsup}): repulsion between classes $u, u'$ down-weighted by $(1 - k(u,u'))$.
\item \textbf{Graded evaluation}: the partial-credit metric of \S\ref{sec:metrics}.
\end{enumerate}

\section{Stage $D$: semi-Markov segmental decoder (textless)}\label{sec:decoder}

The integration layer. A complete labeled segmentation $\sigma$ (in frames: $\sigma_k = (t_k, s_k, e_k)$, $s_{k+1} = e_k$) is scored by
\begin{equation}
S(\sigma) = \sum_{k=1}^{K} \Big[
\underbrace{\lambda_c \sum_{t=s_k}^{e_k-1} \log p_{\mathrm{f}}(t_k \mid \tilde x_t)}_{\text{acoustic evidence (\S\ref{sec:m2f})}}
+ \underbrace{\lambda_d \log P_{\mathrm{dur}}(e_k - s_k \mid t_k)}_{\text{duration prior}}
+ \underbrace{\lambda_b\, \beta_{s_k}}_{\text{boundary evidence (\S\ref{sec:model1})}}
+ \underbrace{\lambda_\ell \log A(t_k \mid t_{k-1})}_{\text{token bigram (optional)}}
\Big].
\label{eq:score}
\end{equation}

\paragraph{Duration model.} Per class, shifted negative binomial $d - 1 \sim \NB(r_u, p_u)$:
\begin{equation}
P_{\mathrm{dur}}(d \mid u) = \binom{d-2+r_u}{d-1} (1-p_u)^{d-1} p_u^{r_u}, \qquad d \ge 1,
\end{equation}
fitted by method of moments from cleaned labels: with $m_u = \bar d_u - 1$, $s_u^2 = \widehat{\mathrm{Var}}(d_u)$, $\;p_u = m_u / s_u^2$, $r_u = m_u^2/(s_u^2 - m_u)$ (fallback to a pooled geometric when $s_u^2 \le m_u$); classes with $< 30$ observations shrink to the pooled estimate. $\sil$ gets a heavy-tailed mixture (short gaps vs.\ instrumental sections): $P_{\mathrm{dur}}(\cdot \mid \sil) = \omega\,\NB(r_1,p_1) + (1-\omega)\,\NB(r_2,p_2)$, EM-fitted.

\paragraph{Token bigram.} $A(u' \mid u)$ is an add-$k$ (or Kneser--Ney) smoothed bigram over the training label sequences. It is a prior over $\Tok$-strings learned offline --- \emph{not} a transcript; decoding remains textless. $\lambda_\ell = 0$ is a valid (pure-acoustic) configuration and is the reporting default until the LM's effect is ablated.

\paragraph{Exact Viterbi DP.} With $D_{\max} = 60$ frames (1.2 s; $\sil$ exempt via chunking at long low-energy runs), cumulative sums $C_u(t) = \sum_{\tau \le t} \log p_{\mathrm{f}}(u \mid \tilde x_\tau)$ make each span score $O(1)$. Define $V(e, u)$ = best score of any segmentation of $[1, e)$ whose last segment has label $u$:
\begin{align}
V(0, u) &= \ind{u = \mathrm{BOS}} \cdot 0, \nonumber\\
V(e, u) &= \max_{1 \le d \le D_{\max}} \max_{u' \in \Tok} \Big[ V(e-d, u') + \lambda_\ell \log A(u \mid u') + \lambda_d \log P_{\mathrm{dur}}(d \mid u) \nonumber\\
&\hspace{9em} + \lambda_c \big(C_u(e-1) - C_u(e-d-1)\big) + \lambda_b\, \beta_{e-d} \Big],
\label{eq:viterbi}\\
S^\star &= \max_u V(T, u), \qquad \sigma^\star \text{ by backpointers}. \nonumber
\end{align}
Complexity $O(T \cdot D_{\max} \cdot |\Tok|^2)$ naive; $O(T \cdot D_{\max} \cdot |\Tok|)$ with $\lambda_\ell = 0$; with a per-span top-$K$ shortlist from $C_u$ ($K = 8$), $O(T \cdot D_{\max} \cdot K^2)$. \emph{Second pass:} the $N$-best segmentations ($N = 4$, extracted by keeping $N$ backpointers) are rescored with the segment LSTM $p_{\mathrm{M2}}$ replacing the frame term, and the best rescored hypothesis is emitted.

\paragraph{Confidence.} The semi-Markov forward--backward (replace $\max$ by $\operatorname{logsumexp}$ in \eqref{eq:viterbi} for $\alpha$; symmetric $\beta$-recursion) yields per-segment posteriors
\begin{equation}
c_k = p\big(\text{segment } (t_k, s_k, e_k) \in \sigma \mid \tilde x_{1:T}\big)
= \frac{\alpha(s_k, \cdot) \cdot [\text{span factors}] \cdot \beta(e_k, t_k)}{Z},
\end{equation}
used for pseudo-label filtering (\S\ref{sec:unsup}) and QA flags (\S\ref{sec:cleaning}).

\paragraph{Degenerations (interpretability).} $\lambda_d = \lambda_\ell = 0$, $\lambda_b \to \infty$ with thresholded $\beta$: recovers the report's two-stage pipeline (Model-1 breaks, then Model-2 per span). $\lambda_c = \lambda_\ell = 0$: pure boundary+duration segmenter. Each term of \eqref{eq:score} is inspectable per decoded segment --- the decoder is a scoring identity, not a black box.

\section{Generative counterpart: sticky HDP-HMM / HSMM (unsupervised)}\label{sec:hdphmm}

Used with \emph{no access to token labels or sequences}: (i) acoustic-unit discovery (\texttt{kashi discover}), (ii) the unsupervised boundary source for dataset cleaning (\S\ref{sec:cleaning}), (iii) posterior boundary uncertainty.

\subsection{Generative model}
Emissions $x'_t = \mathrm{PCA}_{d'}(g_w(x_t)) \in \R^{d'}$, $d' = 48$ (whitened). With truncation $L = 120$ (weak-limit),
\begin{align}
\beta &\sim \Dir(\gamma/L, \dots, \gamma/L), \\
\pi_j &\sim \Dir\big(\alpha\beta_1, \dots, \alpha\beta_j + \kappa, \dots, \alpha\beta_L\big), \qquad j = 1,\dots,L,\\
\theta_j = (\mu_j, \mathrm{diag}(\varsigma_j^2)) &\sim \mathrm{NIG}(\mu_0, \lambda_0, a_0, b_0) \text{ per dimension}, \\
z_t \mid z_{t-1} &\sim \pi_{z_{t-1}}, \qquad x'_t \mid z_t \sim \Norm(\mu_{z_t}, \mathrm{diag}(\varsigma^2_{z_t})).
\end{align}
$\kappa$ is the sticky self-transition mass ($\rho = \kappa/(\alpha+\kappa) \approx 0.95$ expected self-transition, i.e.\ expected dwell $\approx 20$ frames $= 400$ ms $\approx$ mora scale). Boundaries are the \emph{derived} events $\{t : z_t \ne z_{t-1}\}$; no break variable exists, hence no class imbalance exists.

\subsection{Blocked Gibbs sweep}
\begin{enumerate}
\item \textbf{Path (FFBS).} Forward: $\alpha_t(j) \propto \Norm(x'_t \mid \theta_j) \sum_{i} \alpha_{t-1}(i)\, \pi_i(j)$. Backward sampling: $z_T \sim \alpha_T(\cdot)$; $z_t \sim \alpha_t(\cdot)\, \pi_{(\cdot)}(z_{t+1})$. One joint draw of $z_{1:T}$ per sweep.
\item \textbf{Emissions.} Per state $j$, per dimension: with $n_j$ assigned frames, mean $\bar x_j$, scatter $S_j$:
$\lambda_n = \lambda_0 + n_j$, $\mu_n = (\lambda_0\mu_0 + n_j\bar x_j)/\lambda_n$, $a_n = a_0 + n_j/2$, $b_n = b_0 + \tfrac12 S_j + \tfrac{\lambda_0 n_j}{2\lambda_n}(\bar x_j - \mu_0)^2$; draw $\varsigma^2 \sim \mathrm{InvGamma}(a_n, b_n)$, $\mu \sim \Norm(\mu_n, \varsigma^2/\lambda_n)$.
\item \textbf{Transitions.} $\pi_j \sim \Dir(\alpha\beta + \kappa\delta_j + n_{j\cdot})$ with transition counts $n_{jk}$.
\item \textbf{Global weights.} Table counts by the Chinese-restaurant sampler $m_{jk} = \sum_{i=1}^{n_{jk}} \mathrm{Bern}\!\big(\tfrac{\omega_{jk}}{\omega_{jk} + i - 1}\big)$, $\omega_{jk} = \alpha\beta_k + \kappa\ind{j{=}k}$; sticky override correction $\bar m_{jj} = m_{jj} - w_j$, $w_j \sim \mathrm{Binom}\!\big(m_{jj}, \tfrac{\rho}{\rho + \beta_j(1-\rho)}\big)$ (Fox et al.\ 2011); then $\beta \sim \Dir(\gamma/L + \bar m_{\cdot 1}, \dots, \gamma/L + \bar m_{\cdot L})$.
\end{enumerate}
Defaults: $\gamma = 4$, $\alpha = 4$, $\kappa$ set from $\rho = 0.95$, 30 sweeps after 10 burn-in.

\subsection{Boundary posterior and HSMM upgrade}
From $S$ post-burn-in samples $z^{(s)}$, the boundary probability and location dispersion are
\begin{equation}
\hat p_t = \tfrac{1}{S}\textstyle\sum_s \ind{z_t^{(s)} \ne z_{t-1}^{(s)}}, \qquad
\text{and for each matched boundary cluster, } \varsigma^{\mathrm{bnd}}_m = \mathrm{std}_s(\text{position}) \cdot h \text{ [ms]}.
\end{equation}
\emph{HSMM:} replace geometric dwell with $d \sim \NB(r, p)$ per state; the forward message over segment ends becomes
$\alpha_t(j) = \sum_{d=1}^{D_{\max}} \big[\prod_{\tau=t-d+1}^{t} \Norm(x'_\tau \mid \theta_j)\big] P_{\mathrm{dur}}(d \mid j) \sum_{i \ne j} \alpha_{t-d}(i)\, \tilde A(i,j)$,
with backward sampling of $(z, d)$ pairs; complexity $O(T D_{\max} L^2)$, reduced to $O(T D_{\max} L)$ under the sticky/uniform-base factorization. HSMM is adopted iff boundary residuals under the HMM look non-geometric (P6 gate in the roadmap).

\section{Dataset cleaning without forced alignment}\label{sec:cleaning}

Fixes \eqref{eq:jitter} and unlabeled breaths using only \emph{sequence-free, local} operations. The token sequence is never used to infer times (Remark~\ref{rem:nofa}); token identities are only carried along.

\subsection{Candidate boundaries (unsupervised)}
Union of: (i) HDP-HMM/HSMM posterior boundaries $\{(b_m, \hat p_m, \varsigma^{\mathrm{bnd}}_m)\}$ with $\hat p_m \ge 0.3$ (\S\ref{sec:hdphmm}); (ii) spectral-flux onsets: $o_t = \sum_f \max(0, |X_{t,f}| - |X_{t-1,f}|)$, peak-picked after per-band normalization; (iii) voicing transitions: extrema of $|v_t - v_{t-1}|$. Merge within 1 frame; candidate confidence $= \max$ of source scores.

\subsection{Monotone boundary snapping}
Labeled internal boundaries $\hat u_1 < \dots < \hat u_J$ (segment ends, frames); candidates $b_1 < \dots < b_M$. Feasible matches: $|b_m - \hat u_j| \le \Delta_{\max} = 5$ frames (100 ms). Order-preserving partial matching by the alignment DP
\begin{equation}
D(j, m) = \min\big\{ D(j{-}1, m{-}1) + c(j, m),\;\; D(j{-}1, m) + c_{\mathrm{miss}},\;\; D(j, m{-}1) \big\},
\qquad c(j,m) = \frac{|b_m - \hat u_j|}{\Delta_{\max}} - \eta\, \hat p_m,
\end{equation}
with $c_{\mathrm{miss}} = 1.2$, $\eta = 0.5$; matches with $c(j,m) > c_{\mathrm{miss}}$ forbidden. Output per boundary: snapped time $u'_j = b_{m(j)}$ (or unchanged if unmatched), $\texttt{moved\_ms}_j = (u'_j - \hat u_j)h$, uncertainty $\varsigma^{\mathrm{bnd}}_{m(j)}$, and $\texttt{flag}_j \in \{$unmatched, large-shift$\}$. This is matching between two \emph{event sets} in time with moves bounded by 100 ms; no transcript, no token-conditioned path, no global warp --- categorically not forced alignment.

\subsection{Breath/noise tagging and QA flags}
Inside every labeled $\sil$ span, tag maximal runs
\begin{equation}
\noise: \quad \{t : e_t \ge \theta_e \,\wedge\, v_t < \theta_v\} \text{ of length} \ge 3 \text{ frames (60 ms)},
\qquad \theta_e = -35\,\mathrm{dB},\; \theta_v = 0.35
\end{equation}
(energetic but aperiodic $=$ breath/fricative noise; energetic \emph{and} periodic runs are flagged \texttt{missed-vocal} instead --- likely mislabeled lyrics). $\noise$ is a dataset annotation only, never a decoder class; tagged frames are excluded from all training losses. Independent token-level QA (no timing involved): for each snapped segment, if $p_{\mathrm{M2}}(t_k \mid \tilde x_{s'_k:e'_k})$ is in the corpus' bottom $q = 5\%$ for its class, flag \texttt{label-suspect}. Per-song QA gate: quarantine iff mean$|$moved$| > 60$ ms, or $>15\%$ rows flagged, or candidate-recall $< 0.6$. Output: \texttt{clean\_v2} label version + \texttt{realign\_report.csv}; one retrain-and-repeat iteration ($\mathrm{v2} \to \mathrm{v3}$) is accepted only if the gold metrics (\S\ref{sec:metrics}) improve. \emph{Limits (by design):} snapping corrects jitter $\le 100$ ms; wrong/missing tokens and off-script singing are flagged, never silently edited.

\section{Learning from unlabeled audio}\label{sec:unsup}

\subsection{Temporal InfoNCE (fully unsupervised)}
Anchors $a_i = g_w(x_{t_i})$, positives $p_i = g_w(x_{t_i + \delta_i})$, $\delta_i \sim \mathrm{U}\{1,2,3\}$, in-batch negatives, temperature $\tau_c = 0.1$:
\begin{equation}
\mathcal{L}_{\mathrm{nce}} = -\frac{1}{B}\sum_{i=1}^{B} \log \frac{\exp(a_i^\top p_i / \tau_c)}{\sum_{j=1}^{B} \exp(a_i^\top p_j / \tau_c)}.
\end{equation}
Rationale: frames 20--60 ms apart almost surely share phonetic content but differ in local noise/phase $\Rightarrow$ $g_w$ learns nuisance-invariant, content-sensitive similarity. Any encoded audio (labeled or scraped) contributes.

\subsection{Supervised contrastive with soft phonetic negatives}
With token labels $u_i$ on segment-pooled embeddings, positives $P(i) = \{j : u_j = u_i\}$ and kernel-weighted repulsion
\begin{equation}
\mathcal{L}_{\mathrm{sup}} = -\frac{1}{B}\sum_i \frac{1}{|P(i)|}\sum_{p \in P(i)} \log
\frac{\exp(a_i^\top a_p/\tau_c)}{\sum_{j \ne i} \big(1 - k(u_i, u_j)\big) \exp(a_i^\top a_j/\tau_c) + \sum_{p' \in P(i)} \exp(a_i^\top a_{p'}/\tau_c)},
\end{equation}
so \emph{ka} and \emph{ga} ($k = 0.91$) are barely repelled while \emph{ka} and \emph{n} ($k = 0.10$) repel almost fully --- the embedding geometry is asked to mirror phonetic geometry instead of fighting it.

\subsection{Covers (audio-to-audio; later phase)}
For two renditions $a, a''$ of one song: DTW on the cosine-distance matrix of frame features yields a monotone \emph{audio-audio} warp $W^\star = \argmin_W \sum_{(t,t') \in W} \big(1 - \cos(g_w(x_t), g_w(x''_{t'}))\big)$ (Sakoe--Chiba band $\pm 5$ s). Matched frame pairs are positives (same syllable content, different singer/timbre). No transcript is involved; this is signal-to-signal correspondence, orthogonal to Remark~\ref{rem:nofa}.

\subsection{Pseudo-labeling (self-training)}
Decode unlabeled songs with \S\ref{sec:decoder}; accept segments with confidence $c_k \ge \theta_c = 0.9$ into a weak-label pool; retrain $p_{\mathrm{f}}, p_{\mathrm{M2}}$ (and optionally Model 1 from decoded boundaries) on labeled $+$ weak data with weight $w_{\mathrm{weak}} = 0.3$; re-evaluate; stop after two non-improving rounds. This is approximate EM on the decoder's own model, entirely textless.

\subsection{Remark: sequence-marginalized frame training (optional, flagged)}
The frame posterior $p_{\mathrm{f}}$ can alternatively be trained from $(x_{1:T}, t_{1:K})$ pairs by marginalizing over all monotone label-to-frame assignments (CTC-style forward--backward on the \emph{training loss}). This uses the labeled sequence exactly as any supervised loss does, never consumes or produces timestamps, and at inference nothing changes (the decoder of \S\ref{sec:decoder} remains textless). It is strictly an alternative to trusting v2 timestamps during training and remains \emph{off} by default; it is listed because it is the only known way to train $p_{\mathrm{f}}$ that is provably independent of residual timing noise. If the latent-offset loss \eqref{eq:lom} + v2 labels suffice on gold metrics, this option stays off.

\section{Evaluation metrics (formal)}\label{sec:metrics}

Let gold segments $\{(t^\star_k, s^\star_k, e^\star_k)\}$ and predictions $\{(\hat t_k, \hat s_k, \hat e_k)\}$, silences stripped and excluded rows dropped unless stated.
\begin{enumerate}
\item \textbf{Boundary F1@$\tau$}, $\tau \in \{20, 50\}$ ms: greedy monotone 1--1 matching of boundary sets within $\tau$; $F_1 = 2PR/(P{+}R)$. Also mean$|\Delta t|$ over matched pairs.
\item \textbf{SER}: $\mathrm{Lev}\big((\hat t_k)_k, (t^\star_k)_k\big) / K^\star$ (token-level Levenshtein; the end-to-end headline).
\item \textbf{Timed-token F1}: a predicted segment is correct iff $\hat t = t^\star$ for its matched gold segment and $|\hat s - s^\star| \le 50$ ms; matching greedy by start time.
\item \textbf{Partial credit}: $\mathrm{PC} = \E[\, k(\hat t, t^\star) \,]$ over matched pairs (the report's \S7.2 made concrete).
\item \textbf{Noise-span P/R}: prediction and gold $\noise$ spans match iff $\mathrm{IoU} \ge 0.3$.
\end{enumerate}
Model selection uses Boundary F1@50 (Model 1), $\mathcal{L}_{\mathrm{cls}}$/accuracy (Model 2), SER + timed-token F1 (pipeline); promotion gates on the frozen paper test split; gold windows on test songs are reporting-only.

\section{Defaults and complexity}

\begin{center}
\small
\begin{tabular}{@{}llll@{}}
\toprule
Component & Key defaults & Train cost (6 GB GPU) & Inference cost / 4-min song \\
\midrule
Separator $V$ & mel-band roformer ckpt & none (frozen) & $\sim$1--3 min (dominant) \\
Backbone $\Phi$ & JA base ckpt, frozen, 20 ms & none & $\sim$10 s \\
Projection $g_w$ & $768{\to}256{\to}128$, $\tau_c{=}0.1$ & minutes (cached feats) & negligible \\
Voicing $v_t$ & autocorr, $[70,500]$ Hz & --- & $O(T L_w \log L_w)$, seconds \\
Model 1 & $d_m{=}256$, $L{=}2$, $H{=}8$, $W_a{=}100$; $\mathcal{L}_{\mathrm{lom}}$, $\Delta{=}3$ & tens of min & $O(T W_a d_m)$, seconds \\
Model 2 (LSTM) & $n_h{=}144\times2$, dropout $0.5$; $\alpha{=}0.1, p{=}4$ & tens of min & rescoring only \\
Model 2f (frame) & linear/MLP-256 head & minutes & $O(T d\,|\Tok|)$ \\
Decoder $D$ & $D_{\max}{=}60$, $K{=}8$, $N{=}4$; $\lambda_c{=}1$, $\lambda_d{=}0.5$, $\lambda_b{=}0.5$, $\lambda_\ell{\in}\{0, 0.3\}$ & $\lambda$'s by grid on dev & $O(T D_{\max} K^2)$, seconds \\
HDP-HMM & $L{=}120$, $d'{=}48$, $\rho{=}0.95$, 30+10 sweeps & --- (CPU ok) & $O(S\,T L D_{\max})$ HSMM, minutes \\
Snapping & $\Delta_{\max}{=}5$ fr, $c_{\mathrm{miss}}{=}1.2$, $\eta{=}0.5$ & --- & $O(JM)$ DP, ms \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.5em}
\noindent\textbf{Two-stage compatibility.} The report's pipeline is the decoder degeneration $\lambda_d = \lambda_\ell = 0$ with hard Model-1 boundaries; it is retained as a registry configuration and as the P1 baseline. All weights $\lambda$ and thresholds above are config keys, not code constants.

\end{document}
